\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move

### 2025-06-10 writing/manuscript.Rmd
addressed some of KH's comments
continue at "Effects of temperature on habitat selection"

---
title: "\\Large Predicting how climate change will affect how and where terrestrial mammals will move"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: TJ Gooliaff
    institute: [forest]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'global-change-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs} # to prevent tables from splitting
subparagraph: true # needed for \usepackage[small]{titlesec}
urlcolor: cyan
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change. Consequently, it is still unclear how and where mammals will move in the 21^st^ century. We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada. We show that temperature is a strong determinant of when, how, and where mammals move, and that climate change will impact mammals' movement rates and habitat selection throughout the 21^st^ century. We found no common effects of temperature on the species’ movement rates, so we suggest that species' estimated responses be interpreted jointly when making inferences about how climate change will impact ecological communities. In contrast, habitat selection strength decreased for all species within their current ranges, suggesting that all six species will shift their range in the coming decades. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation. We thus conclude by providing suggestions for conservation and future research while calling for a more holistic and collaborative framework."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

\noindent \textbf{Figures}: 7

\noindent \textbf{Tables}: 2

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [e.g., @hall_microrefuges_2016]. Additionally, temperature also affects movement indirectly through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010]. Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

Temperature also affects the consequences of animal movement, such as encounter rates with resources [@mezzini_how_2025], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023]. But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes outside their typical range [e.g., @giroux_activity_2023], while thermal generalists can adapt more easily to change [@dupont_beyond_2024]. However, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to respond within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].

@@

In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@ipcc_climate_2023; @zurowski_summer_2023]. Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. Consequently, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Uncertainty in future conditions and their consequences on ecosystems present a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behaviour and (2) the implications this will have on populations and species' adaptability and resiliency, community structure, and biodiversity at large. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@ipcc_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].

@@

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to be able to prepare for and respond to change in a proactive manner [e.g., @sawyer_identifying_2009; @mccain_assessing_2019]. The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [e.g., @tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [e.g., @zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying which landscapes are most valuable now as well as in the future. Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, pumas: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data thoughout a large spatial range of British Columbia and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSPs -- see @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and landscape levels.

@@

This study leverages four main datasets: (1) a multi-species collection of GPS telemetry data (Fig. \@ref(fig:bc-map)); (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017]. We detail the data sources and analyses in the sections below.<!--, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).-->

@@

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)). ***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, a total of 2396 GPS locations (0.16%, including collar calibration data) were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour. See Appendix A for additional information.

### Historical and projected temperature data

\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].

We obtained rasters of projected monthly average temperature at a 0.08$^\circ$ resolution in BC from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, see @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year. We then modeled the estimated variance using a GAM for location and scale [GAMLS, see: @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month. The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space. We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.<!-- Additional details are available in Appendix C.-->

@@

We simulated hourly variation in future weather by assuming hourly temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS. We then predicted changes in movement behaviour and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs. For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile<!-- (Appendix C)-->.

@@

\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`. We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019]. The 18 animals with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).

Since `ctmm``s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates, as $k$-nearest neighbors did not discriminate between states well (Fig. B3).

@@

\noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement. To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models. We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. B1-B2).

@@

\noindent In this study, we separate the effects of temperature on mammalian movement into indirect and direct effects. We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct". For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behaviour by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly). We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. B1-B2).

>>>>

---
title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: TJ Gooliaff
    institute: [forest]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'global-change-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs} # to prevent tables from splitting
subparagraph: true # needed for \usepackage[small]{titlesec}
urlcolor: cyan
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

```{r abstract, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have quantitatively estimated the effects of temperature on mammalian movement behaviour, but few have estimated the effects of future climate change. Consequently, it is still unclear how and where mammals will move in the 21^st^ century. We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada, between 1998 and 2023. We show that temperature is a strong determinant of when, how, and where mammals move, and we predict that climate change will impact mammals' movement rates and habitat selection as temperatures warm throughout the 21^st^ century. We found the effects of temperature on movement rates varied across species, so we suggest that species' estimated responses not be interpreted individually when making inferences about how climate change will impact ecological communities. In contrast, habitat selection strength decreased for all species within their observed ranges, suggesting that all six species may shift their range in the coming decades. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation. We thus conclude by providing suggestions for conservation and future research while calling for a more holistic and collaborative framework."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

\noindent \textbf{Figures}: 8

\noindent \textbf{Tables}: 3

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [e.g., @hall_microrefuges_2016]. Additionally, temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and novel phenology [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022]. Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

<!-- since movement depends on temperature, the consequences of movement do, too -->

By impacting animal movement behavior, temperature also affects the consequences of movement, such as encounter rates with resources [@mezzini_how_2025], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023]. But what counts as "extreme" varies across species: Species that experience narrow temperature ranges tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while thermal generalists can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. However, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to respond within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].

@@

In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023]. Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. Consequently, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Uncertainty in future conditions and their consequences on ecosystems present a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behaviour and (2) the implications such changes and uncertainty will have on populations and species' adaptability and resiliency, as well as overall community structure. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [e.g., body size, activity time, and elevational and latitudinal distribution @mccain_assessing_2019]. However, it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@ipcc_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].

@@

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to respond to change in a proactive manner [@sawyer_identifying_2009; @mccain_assessing_2019]. The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying landscape dynamics and value will evolve throughout the century. Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

This paper provides a framework for quantifying the effects of proximal air temperature on animal movement using quantitative methods that are flexible and statistically robust. We present results for six large-bodied terrestrial mammal species in British Columbia (BC), Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of British Columbia and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and community levels.

@@

This study leverages four main datasets: (1) a multi-species collection of GPS telemetry data (Fig. \@ref(fig:bc-map)); (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017]. While we acknowledge that forest type, structure, and age are important variables, we do not include such variables in this study for the sake of parsimony and comparability across species. We detail the data sources and analyses in the sections below.

@@

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)). ***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, a total of 2396 GPS locations (0.16%, including collar calibration data) were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour (Appendix A).

### Historical temperature data and temperature projections

\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 database [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].

We obtained rasters of projected monthly average temperature at a 0.08$^\circ$ resolution in BC from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year. We then modeled the estimated variance using a GAM for location and scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month. The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space. We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.

@@

We simulated hourly variation in future weather by assuming hourly temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS. For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile.

@@

\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`. We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, @fleming_fine-scale_2014; @noonan_scale-insensitive_2019]. The 18 animals with insufficiently fine sampling to support movement models with finite speed [@noonan_scale-insensitive_2019] were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).

Since `ctmm``s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using a $k$-means algorithm with 2 clusters for each species' detrended speeds. When the algorithm clearly failed to discriminate between states, we estimated the split point using the inflection points in histograms of the detrended speeds (Fig. B1).

@@

\noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement. To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models.

@@

\noindent In this study, we separate the effects of temperature on mammalian movement into direct and indirect effects. We call "direct" all effects that impact movement behavior causally without first impacting another variable (Fig. \@ref(fig:dags)). For example, in Fig. \@ref(fig:dags)A, whether a grizzly bear is moving in a given moment directly depends on time of day and day of year, since it will not be moving when sleeping or hibernating. However, time of day and day of year also impact its movement behavior indirectly because its sensitivity to temperature will also depend on time of day and day of year. Thus, both time of day and day of year also impact movement frequency indirectly by altering the effect of temperature on movement frequency. Failing to disentangle the effects of these three variables will lead to incorrect and likely confusing estimates for the causal effect of temperature on movement behavior, since causal and non-causal relationships would be indistinguishable [@mcelreath_statistical_2020].

Indirect effects of temperature also include all effects that can be controlled by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022]. For example, in Fig. \@ref(fig:dags)B, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can mitigate these indirect effects of temperature by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022]. In this case, it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly).

```{r dags,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graphs (DAGs) assumed for inferring the causal effects of temperature (Temp) on movement behavior. (A) DAG for the probability of moving, speed when moving, or distance traveled ($Y$) while accounting for the effects of time of day (TOD), day of year (DOY), and their interaction effects. Temperature directly affects $Y$, but the effects of temperature depend on the time of day and season. Time of day and day of year also affect $Y$ directly, but the effect of time of day changes throughout the year due to changes in day length and seasonality. (B) DAG for Relative Selection Strength (RSS) for percent forest cover (F), elevation (E), and distance from water (W). The RSS for of a given habitat depends on all three resources, and the selection for each resource is independent of the other two resources and dependent on temperature.", fig.height=4, fig.width=9, out.width='\\linewidth'}
library('dplyr')   # for data wrangling
library('ggplot2') # for fancy plots
library('ggdag')   # for directed acyclical graphs
library('cowplot') # for plots in grids
